Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-05-06 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
minor changes + new text regarding Carullo's

%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
% General:
\redsout{The detection of gravitational waves produced by coalescing compact binaries
allows}
\hs{Detections of gravitational waves from coalescing compact binaries allow}
us to explore \hs{into} the dynamical, nonlinear regime of general relativity
and \redsout{to} constrain modifications to it.
%
Some of these gravitational-wave events have sufficiently high
signal-to-noise ratio in the merger, allowing us to probe the relaxation of the remnant black
hole to its final, stationary state --- the so-called black-hole ringdown, which is
characterized by a set of quasinormal modes.
% Main question:
Can we use the ringdown to constrain deviations \redsout{to} \hs{from} general relativity,
as predicted by several of its contenders?
% What we do:
Here, we  address this question by using an inspiral-merger-ringdown waveform
model in the effective-one-body formalism, augmented with a parametrization of
the ringdown based on an expansion in the final black hole's spin.
%
We give a prescription on how to include in this waveform model\hs{,} the quasinormal mode frequencies
calculated on a theory-by-theory basis. In particular, we focus on theories that modify
general relativity by higher-order curvature corrections, namely,
Einstein-dilaton-Gauss-Bonnet\hs{,} \redsout{and} dynamical Chern-Simons theories, and cubic- and quartic-order
effective-field-theories of general relativity.
%
% Main conclusion:
\redsout{We confront the resultant waveform models against the two loudest}
\hs{We use this parametrized waveform model to measure the ringdown properties of}
the two loudests \redsout{gravitational-wave} ringdown signals observed so far, \redsout{notably,} GW150914 and GW200129.
%
We find that while the Einstein-dilaton-Gauss-Bonnet theory cannot be
constrained with these events, we can place upper bounds on the
fundamental length-scale of cubic- ($\ell_{\rm cEFT} \leqslant 38.2$~km) and quartic-order
($\ell_{\rm qEFT} \leqslant 51.3$~km) effective-field-theories of general
relativity, and of dynamical Chern-Simons gravity ($\ell_{\rm dCS} \leqslant 38.7$~km).
%
The latter result is a concrete example of a theory presently unconstrained by
inspiral-only analyses which, however, can be constrained by merger-ringdown
studies with current gravitational-wave data.
%
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%

@@

Several null tests have been proposed to probe the nature of gravity with GW signals~\cite{TheLIGOScientific:2016src,Yunes:2016jcc,Abbott:2018lct,LIGOScientific:2019fpa,Abbott:2020jks,LIGOScientific:2021sio}. They include tests of GW generation~\cite{Arun:2006yw,Yunes:2009ke,Li:2011cg,Agathos:2013upa,Mehta:2022pcn}, where deviations in the post-Newtonian (PN) coefficients in the inspiral, and phenomenological coefficients in the plunge and merger stages can be bounded; tests of GW propagation~\cite{Will:1997bb}, which allow us to set upper limits on coefficients entering generalized dispersion relations, including the Compton wavelength associated to the
mass of the graviton; tests of the polarization of gravitational radiation~\cite{Will:2014kxa}, for which more than
two GW detectors are needed to set statistically significant bounds, and tests of the remnant properties~\cite{Meidam:2014jpa,Carullo:2018sfu,Brito:2018rfr,Carullo:2019flw,Isi:2019aib,Ghosh:2021mrv,Carullo:2021dui}
\redsout{during} \hs{in the post-merger}. So far, none of these null tests have reported any deviation from GR.

@@

\begin{table}[t]
\begin{threeparttable}
\begin{tabular}{c | c c}
\hline \hline
Theory & Constraint & This work \\
\hline
EdGB        & $\ell_{\rm EdGB} \leqslant 1.18$~km~(GW)~\cite{Lyu:2022gdr} & -- \\
dCS         & $\ell_{\rm dCS} \leqslant 8.5$~km~(EM+GW)~\cite{Silva:2020acr}  & $\ell_{\rm dCS} \leqslant 38.7$~km \\
cubic EFT   & -- & $\ell_{\rm cEFT} \leqslant 38.2$~km \\
quartic EFT & $\ell_{\rm qEFT} \leqslant 150$~km~\cite{Sennett:2019bpc}  & $\ell_{\rm qEFT} \leqslant 51.3$~km~\tnote{a}
\\
\hline \hline
\end{tabular}
\caption{Summary of the upper limits, at $90\%$ credible level,
on the length-scale of the modified gravity theories under investigation.
They were obtained combining Bayesian-inference results from GW150914 and GW200129.
Current constraints on $\ell_{\rm th}$ are also listed.}
\begin{tablenotes}
\item[a]{\footnotesize We notice that our result for the quartic EFT of GR is only in marginal tension with our hard cut-off scale for the
validity of the theory, and hence we do still quote it here (see Secs.~\ref{sec:remarks} and~\ref{sec:results_qeft}
for details)\hs{.}}
\end{tablenotes}
\label{tab:bound_summary}
\end{threeparttable}
\end{table}

@@

In the end, the $\pSEOB$ waveform model with theory-specific
QNMs has only $\ell_{\rm th}$ as a free beyond-GR parameter.
%
We emphasize that our procedure is different from that of
Ref.~\cite{Carullo:2021dui} which, for a given value of $p$, varied all $\ell_{\rm th}$,
$\delta\omega^{(j)}$, and $\delta\tau^{(j)}$ parameters, and then proceeded
to use the posteriors on $\ell_{\rm th}$
\hs{, considering up to $j=2$ in
the GR deformation coefficients, and remaining agnostic about the underlying
theory which would predict the modifications to the QNMs.}
%
\redsout{to draw conclusions to any theory that predict\hs{s}
corrections at a given $p$.}
%
\redsout{This is a important difference between our work and Ref.~\cite{Carullo:2021dui}.
In particular, as we will see in Sec.~\ref{sec:results},}
\hs{We will see in Sec.~\ref{sec:results} that} adding theory-specific
information to the ParSpec coefficients \emph{can lead to different interpretations
of the bounds on $\ell_{\rm th}$, even for different theories that predict the same
value of the exponent $p$.}

@@

For $N$ independent GW observations $\{d_j\}$, $j=1,...,N$, each characterized
by a PDF $P_j(\ell_{\rm th} \vert d_j,\mathcal{H})$, the
joint posterior can be written as:
%
\begin{equation}
P(\ell_{\rm th} | \{d_j\},\mathcal{H}) = p(\ell_{\rm th}) \prod_{j=1}^{N} \frac{P_j(\ell_{\rm th} | d_j,\mathcal{H})}{p_j(\ell_{\rm th}|\mathcal{H})}\,.
\label{eq:cumulative_dist_ell}
\end{equation}
%
where $p_j(\ell_{\rm th} | \mathcal{H})$ are the priors used for each observation, \redsout{and} $p(\ell_{\rm th} | \mathcal{H})$ is an overall prior\hs{, and we assume that the value of $\ell_{\rm th}$ is shared among all events.}
%
Since we assume a uniform prior on $\ell_{\rm th}$, the joint posterior is
equal to the joint likelihood. Hereafter, we will drop the explicit usage of
$\mathcal{H}$.

@@

In Sec.~\ref{sec:results}, we enumerate through the different theories and
outline the main results. Whenever possible, we also combine results from
\redsout{multiple} \hs{both} events to obtain the strongest possible bound on $\ell_{\rm th}$.

@@

We can compare this conclusion with that of Ref.~\cite{Carullo:2021dui}, which obtained $\ell_{\rm EdGB} \lesssim 35$~km,
by setting $p=4$, \emph{but not} including theory-specific QNM information on $\delta \omega^{(j)}$ and $\delta \tau^{(j)}$.
%
Our results provide a concrete example of the importance of including
theory-specific QNM calculations information into the parameter estimation
and how this can dramatically change the outcome of the results.

@@

% --------------------------
\section*{Acknowledgements}
\label{sec:acknowledgements}
%
We thank Emanuele Berti, Andrea Maselli, Caio F. B. Macedo, Deyan Mihaylov,
Serguei Ossokine, Scott E. Perkins, and Helvi Witek for discussions.
%
\hs{We also thank Anuradha Gupta and Gregorio Carullo for comments on this work.}
%
We are grateful for the computational resources provided by the Max Planck
Institute for Gravitational Physics in Potsdam, specifically the
high-performace computing cluster Hypatia and to Steffen Grunewald for
assistance.
%
The material presented in this manuscript is based upon work supported by NSF’s LIGO Laboratory,
which is a major facility fully funded by the NSF.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
% General:
Detecting gravitational waves from coalescing compact binaries allows  
us to explore the dynamical, nonlinear regime of general relativity
and constrain modifications to it.
%
Some of the gravitational-wave events observed by the LIGO-Virgo Collaboration have sufficiently high
signal-to-noise ratio in the merger, allowing us to probe the relaxation of the remnant black
hole to its final, stationary state --- the so-called black-hole ringdown, which is
characterized by a set of quasinormal modes.
% Main question:
Can we use the ringdown to constrain deviations from general relativity,
as predicted by several of its contenders?
% What we do:
Here, we  address this question by using an inspiral-merger-ringdown waveform
model in the effective-one-body formalism, augmented with a parametrization of
the ringdown based on an expansion in the final black hole's spin.
%
We give a prescription on how to include in this waveform model, the quasinormal mode frequencies
calculated on a theory-by-theory basis. In particular, we focus on theories that modify
general relativity by higher-order curvature corrections, namely,
Einstein-dilaton-Gauss-Bonnet, dynamical Chern-Simons theories, and cubic- and quartic-order
effective-field-theories of general relativity.
%
% Main conclusion:
We use this parametrized waveform model to measure the ringdown properties of 
the two loudests ringdown signals observed so far, GW150914 and GW200129.
%
We find that while the Einstein-dilaton-Gauss-Bonnet theory cannot be
constrained with these events, we can place upper bounds on the
fundamental length-scale of cubic- ($\ell_{\rm cEFT} \leqslant 38.2$~km) and quartic-order
($\ell_{\rm qEFT} \leqslant 51.3$~km) effective-field-theories of general
relativity, and of dynamical Chern-Simons gravity ($\ell_{\rm dCS} \leqslant 38.7$~km).
%
The latter result is a concrete example of a theory presently unconstrained by
inspiral-only analyses which, however, can be constrained by merger-ringdown
studies with current gravitational-wave data.
%
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%

@@

Several null tests have been proposed to probe the nature of gravity with GW signals~\cite{TheLIGOScientific:2016src,Yunes:2016jcc,Abbott:2018lct,LIGOScientific:2019fpa,Abbott:2020jks,LIGOScientific:2021sio}. They include tests of GW generation~\cite{Arun:2006yw,Yunes:2009ke,Li:2011cg,Agathos:2013upa,Mehta:2022pcn}, where deviations in the post-Newtonian (PN) coefficients in the inspiral, and phenomenological coefficients in the plunge and merger stages can be bounded; tests of GW propagation~\cite{Will:1997bb}, which allow us to set upper limits on coefficients entering generalized dispersion relations, including the Compton wavelength associated to the
mass of the graviton; tests of the polarization of gravitational radiation~\cite{Will:2014kxa}, for which more than
two GW detectors are needed to set statistically significant bounds, and tests of the remnant properties~\cite{Meidam:2014jpa,Carullo:2018sfu,Brito:2018rfr,Carullo:2019flw,Isi:2019aib,Ghosh:2021mrv,Carullo:2021dui} in the post-merger stage. So far, none of these null tests have reported any deviation from GR.

@@

\begin{table}[t]
\begin{threeparttable}
\begin{tabular}{c | c c}
\hline \hline
Theory & Constraint & This work \\
\hline
EdGB        & $\ell_{\rm EdGB} \leqslant 1.18$~km~(GW)~\cite{Lyu:2022gdr} & -- \\
dCS         & $\ell_{\rm dCS} \leqslant 8.5$~km~(EM+GW)~\cite{Silva:2020acr}  & $\ell_{\rm dCS} \leqslant 38.7$~km \\
cubic EFT   & -- & $\ell_{\rm cEFT} \leqslant 38.2$~km \\
quartic EFT & $\ell_{\rm qEFT} \leqslant 150$~km~\cite{Sennett:2019bpc}  & $\ell_{\rm qEFT} \leqslant 51.3$~km~\tnote{a}
\\
\hline \hline
\end{tabular}
\caption{Summary of the upper limits, at $90\%$ credible level,
on the length-scale of the modified gravity theories under investigation.
They were obtained combining Bayesian-inference results from GW150914 and GW200129.
Current constraints on $\ell_{\rm th}$ are also listed.}
\begin{tablenotes}
\item[a]{\footnotesize We notice that our result for the quartic EFT of GR is only in marginal tension with our hard cut-off scale for the
validity of the theory, and hence we do still quote it here (see Secs.~\ref{sec:remarks} and~\ref{sec:results_qeft}
for details).}
\end{tablenotes}
\label{tab:bound_summary}
\end{threeparttable}
\end{table}

@@

In the end, the $\pSEOB$ waveform model with theory-specific
QNMs has only $\ell_{\rm th}$ as a free beyond-GR parameter.
%
We emphasize that our procedure is different from that of
Ref.~\cite{Carullo:2021dui} which, for a given value of $p$, varied all $\ell_{\rm th}$,
$\delta\omega^{(j)}$, and $\delta\tau^{(j)}$ parameters, and then proceeded
to use the posteriors on $\ell_{\rm th}$
\hs{, considering up to $j=2$ in
the GR deformation coefficients, and remaining agnostic about the underlying
theory which would predict the modifications to the QNMs.}
%
\hs{We will see in Sec.~\ref{sec:results} that} adding theory-specific
information to the ParSpec coefficients \emph{can lead to different interpretations
of the bounds on $\ell_{\rm th}$, even for different theories that predict the same
value of the exponent $p$.}

@@

For $N$ independent GW observations $\{d_j\}$, $j=1,...,N$, each characterized
by a PDF $P_j(\ell_{\rm th} \vert d_j,\mathcal{H})$, the
joint posterior can be written as:
%
\begin{equation}
P(\ell_{\rm th} | \{d_j\},\mathcal{H}) = p(\ell_{\rm th}) \prod_{j=1}^{N} \frac{P_j(\ell_{\rm th} | d_j,\mathcal{H})}{p_j(\ell_{\rm th}|\mathcal{H})}\,.
\label{eq:cumulative_dist_ell}
\end{equation}
%
where $p_j(\ell_{\rm th} | \mathcal{H})$ are the priors used for each observation, \redsout{and} $p(\ell_{\rm th} | \mathcal{H})$ is an overall prior, and we assume that the value of $\ell_{\rm th}$ is shared among all events.
%
Since we assume a uniform prior on $\ell_{\rm th}$, the joint posterior is
equal to the joint likelihood. Hereafter, we will drop the explicit usage of
$\mathcal{H}$.

@@

In Sec.~\ref{sec:results}, we enumerate through the different theories and
outline the main results. Whenever possible, we also combine results from 
both events to obtain the strongest possible bound on $\ell_{\rm th}$.

@@

We can compare this conclusion with that of Ref.~\cite{Carullo:2021dui}, which obtained $\ell_{\rm EdGB} \lesssim 35$~km,
by setting $p=4$, \emph{but not} including theory-specific QNM information on $\delta \omega^{(j)}$ and $\delta \tau^{(j)}$. 
\ab{Furthermore, Ref.~\cite{Carullo:2021dui} did not impose the EFT and ParSpec bounds that we imposed.} 
%
Our results provide a concrete example of the importance of including
theory-specific QNM calculations information into the parameter estimation
and how this can dramatically change the outcome of the results.

@@

% --------------------------
\section*{Acknowledgements}
\label{sec:acknowledgements}
%
We thank Emanuele Berti, Andrea Maselli, Caio F. B. Macedo, Deyan Mihaylov,
Serguei Ossokine, Scott E. Perkins, and Helvi Witek for discussions.
%
We also thank Anuradha Gupta and Gregorio Carullo for comments on this manuscript. 
%
We are grateful for the computational resources provided by the Max Planck
Institute for Gravitational Physics in Potsdam, specifically the
high-performace computing cluster Hypatia and to Steffen Grunewald for
assistance.
%
The material presented in this manuscript is based upon work supported by NSF’s LIGO Laboratory,
which is a major facility fully funded by the NSF.
